\Ace: An Actively-Typed Compilation Environment\\for High-Performance Computing

### 2012-09-14 ipdps13/ace-ipdps13.tex
Renamed clq to ace in file names.

@@

\begin{abstract}
In this paper, we introduce the Ace compilation environment and demonstrate its practicality as a foundational tool in both high-performance computing research and in practice. Ace consists of a statically-typed language that shares syntax with Python and uses the Python language itself as a multi-purpose compile-time metalanguage. Unlike standard languages, however, the Ace compiler consists only of a minimal core that delegates control over most aspects of compilation to user libraries using a novel extension mechanism that we call {\em active type-checking and translation (\ATT)}. In particular, users can specify new primitive types and operations by equipping type definitions, which are first-class objects in the metalanguage, with special methods. The compiler then invokes these methods when type checking and translating expressions, according to a fixed type dispatch protocol associated with each syntactic form. Translation targets a user-defined backend language, and the safety of user-defined translations can be checked automatically. Using this mechanism, we implement the primitives of the C99 and OpenCL languages as simple modules within Ace. We go on show examples of other interesting primitive parallel  abstractions that build on top of these basic modules, demonstrating the foundational role that an actively-typed system like Ace can play in support of research on language-based parallel programming abstractions.

In addition to \ATT, Ace supports more conventional forms of structured and unstructured metaprogramming  and integrates directly with the numeric capabilities of Python itself. We use these features to design a scientific simulation framework that allows users to modularly specify and orchestrate the execution of parameterized families of simulations on clusters of GPUs. This framework is used to successfully conduct large-scale, high-performance neuroscience simulations, providing  evidence that Ace can be used in practice today.

\end{abstract}

@@

Computer-aided simulation and data analysis techniques have transformed science and engineering. Surveys show that scientists and engineers now spend up to 40\% of their time writing software, generally individually or in small groups \cite{howison2011scientific}\cite{hannay2009scientists}. Most of this software targets desktop hardware, but about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}. To fully harness the power of modern computing resources, however, these so-called {\em professional end-user developers} \cite{segal2007some} must write parallel programs.

Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, and statically-typed, low-level languages like C and Fortran paired with low-level abstractions like pthreads and MPI for performance-critical tasks. These low-level languages and abstractions are notoriously tedious and difficult to use correctly, even for expert programmers. CITATION?

>>>>

\usepackage{color}

@@

\begin{abstract}
In this paper, we introduce the Ace compilation environment and demonstrate its practicality as a foundational tool for both high-performance computing research and in practice. 
Ace consists of a statically-typed language that shares syntax with Python and uses the Python language itself as a multi-purpose compile-time metalanguage. Unlike standard languages, however, the Ace compiler consists only of a minimal core that delegates control over most aspects of front-end compilation to {\em active libraries} using a novel extension mechanism that we call {\em active type-checking and translation (\ATT)}. Users specify new primitive types and operations by equipping type definitions, which are first-class objects in the metalanguage, with designated methods that the compiler selectively invokes when type checking and translating expressions. Using \ATT, we implement the primitives of the C99 and OpenCL languages as modules within Ace. We then give examples of several first-class parallel abstractions that can be used alongside these basic primitives, demonstrating the foundational role that an actively-typed system like Ace can play for researchers developing language-based programming abstractions.

Ace also supports more conventional forms of structured and unstructured metaprogramming and Ace programs can be compiled and launched from Python directly. We used these features, along with the OpenCL language module, to design a scientific simulation framework that allows users to modularly specify and orchestrate the execution of parameterized families of simulations on clusters of GPUs. This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing  evidence that Ace is also useful in practice today.
\end{abstract}

@@

Computer-aided simulation and data analysis techniques have transformed science and engineering. Surveys show that scientists and engineers now spend up to 40\% of their time writing software, generally individually or in small groups \cite{howison2011scientific}\cite{hannay2009scientists}. Most of this software targets desktop hardware, and about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}. To fully harness the power of modern processors, however, these so-called {\em professional end-user developers} \cite{segal2007some} must write parallel programs.

Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as simple data analysis and plotting \cite{nguyen2010survey}. For tasks where these interpretted, dynamically-typed languages are too slow, they turn to statically-typed, low-level languages like C and Fortran, paired with low-level abstractions like pthreads and MPI \cite{4222616}\cite{basili2008understanding}. These low-level languages and abstractions are notoriously tedious and difficult to use correctly, even for expert programmers. {\color{red} CITATION?}
